Functional traits complicate the picture of temporal biodiversity change in bird and mammal communities

### 2024-04-01 paper/GEB Submission/manuscript_GEB.Rmd
Reformat discussion

`r colorize("The expectation for functional change across communities is not obvious from past work, and may or may not follow taxonomic trends", "blue")` [@dornelas2023].
While `r colorize("loss of functional diversity", "blue")` is frequently cited as one of the most pressing concerns of the anthropocene [@cardinale2012; @dirzo2014; @young2016], `r colorize("functional diversity may be maintained even when species are lost from a community", "blue")` [@daz2001; @petchey2009].
Forecasts of functional loss range from negligible [@gallagher2013] to dire [@petchey2002; @pimiento2020].
And while some observed trends show significant functional loss [@flynn2009] others document no loss even in some of the most heavily impacted communities [@edwards2013; @larsen2018; @matuoka2020].
On paleoecological time scales, `r colorize("the functional space", "blue")` shows mixed responses to environmental change and extinction events [@jackson2015; @dornelas2018], with significant impacts of species extinctions on functional diversity in some taxa and not others [@pimiento2017].
For some time periods, functional diversity appears to be maintained for substantial portions of geological time [@hedberg2022].
`r colorize("Contemporary, broad-scale examinations of functional change are limited to only a few taxa-focused studies, but show for example functional richness increases for both North American birds", "blue")` [@barnagaud2017; @jarzyna2016] `r colorize("and ray-finned fishes, sharks, and rays", "blue")` [@trindade-santos2020].

Here we `r colorize("leverage ongoing efforts to assemble functional trait data and recent computational advances to", "blue")` perform the first multi-taxa, multi-realm assessment of functional diversity change through time.
We focus on mammal and bird species as `r colorize("a subset of the world's biodiversity of particular conservation concern that is heavily impacted by anthropogenic change.", "blue")`While examining trends in plants, invertebrates, and other vertebrate species is of equal interest, trait data for those taxa raise additional challenges such as limited and biased species coverage [@fitzjohn2014], a lack of accepted species-level means, and differences in the types of traits collected.
To ensure comparability across taxa in trait type and data quality we therefore focus on mammals and birds.
`r colorize("We include body mass, dietary, foraging and other behavioral traits", "blue")` that were intentionally selected to be representative of a species' Eltonian niche, thereby summarizing the functional role they play in the community [@wilman2014].
An initial assessment of amphibian trends is included in the supplement, but excluded from general trend assessment here due to limited geographic coverage.

We assess thousands of mammal and bird functional diversity time series to determine whether or not `r colorize("the addition of functional trait data gives a clarifying picture of biodiversity change across communities. Rather than testing specific hypotheses of change we present a few areas of change consensus across communities, and even more scenarios where the relationship between functional and species change are unexplained and warrant further theoretical and experimental examination. We present results for three different levels: general trends across communities, trends for communities with similar characteristics (taxa, biome, realm, protection status), and trends for individual studies.", "blue")`

@@

```{r traitTable, echo=FALSE}
tibble(Trait = c(rep("Diet", 10), rep("Foraging Strata", 7), rep("Pelagic Specialist", 2), rep("Nocturnal", 2), "Body Mass"),
       Category = c("Invertebrate", "Mammals and Birds", "Reptiles", "Fish", "Unknown Vertebrates", "Scavenging", "Fruit", "Nectar", "Seeds", "Other Plant", 
                    "Below water surface", "water surface", "ground", "understory", "> 2m, below canopy", "canopy", "aerial",
                    "yes", "no",
                    "yes", "no",
                    "-"),
       `Data Type` = c(rep("percentage consumed", 10), rep("percentage of use", 7), rep("binary", 4), "continous, in grams")) %>%
  kable(format = "latex", caption = "Description of the traits included in the analysis broken down by categories at data type.") %>%
  collapse_rows(columns = 1:3) %>%
  kable_paper()
```

@@

Functional diversity metrics were calculated using the *dbFD* function from the *FD* R package [@laliberté2010].
Here we report functional richness (*FRic*), functional evenness (*FEve*), and functional divergence (*FDiv*) which together describe three complementary characteristics of the functional space [@mason2005; @hillebrand2009].
*FRic* assesses the volume of the trait space occupied by species in the community, with higher values indicating communities with species of more extreme trait values.
*FEve* describes how species are distributed across the trait space and how abundance is distributed across species.
Higher values of *FEve* indicate more even spacing of species in the trait space and individuals across species.
*FDiv* measures the degree to which species and their abundances maximize differences in the functional space.
Higher values of *FDiv* therefore correspond to communities where many highly abundant species are on the edges of the trait space.
We also calculated the community-weighted mean (*CWM*) of included traits to examine shifts in the distribution of each trait.
`r colorize("Hereafter we refer to results for functional diversity metrics (FRic, FEve, FDiv) and composition metrics (trait CWM's).", "blue")`

@@

### Evidence of consensus

@@

`r colorize("We found evidence of functional richness loss in mammal studies, with a loss trend significantly different from zero and significantly different from the trend for birds, indicating a potential loss of functional capacity in these communities. This result is consistent with previous work linking anthropogenic drivers such as habitat loss, poaching, and human modification and development to loss of functional diversity in mammal communities across the globe", "blue")` [@brodie2021; @cox2022; @li2022].
`r colorize("However we did not find trends outside the random expectation for species loss, in contrast to expectations for multiple continents based on IUCN risk categories", "blue")` [@brodie2021].
`r colorize("In general, functional capacity of mammals shows greater declines that bird communities.", "blue")`

`r colorize("We also found evidence of increasing evenness for Marine studies. The empirical link between changes in evenness and ecosystem process is the most poorly studied of the metrics, however theory indicates that increased evenness improves function and stability ", "blue")` [@filstrup2019], `r colorize("as the dominance of one or a few species makes function sensitive to the ability of those species to respond to environmental change ", "blue")`[@hillebrand2008; @hillebrand2009].
`r colorize("Further, there is some evidence that maintaining evenness is important for supporting multifunctionality in communities ", "blue")`[@bagousse-pinguet2019].
`r colorize("The majority of marine studies are of seabird communities (7 of 9), so the general trend may be an early indicator of pay off from investments in sea bird recovery, which is also indicated in recent species-based work ", "blue")`[@spatz2023].

`r colorize("Compositional shifts shown by significant trends in CWM's were restricted to dietary traits, with no evidence of a trend in foraging traits or body mass. Increases in fish consumption for bird and temperate studies my be the product of increased availability of easy fish from commercial fishing operations as fish are processed or stocking of inland waterways for sport fishing. However, fisheries also have negative impacts on waterbirds through competition for fish or bird mortality due to by catch. Decreases in nectar consumption for tropical studies is likely due declines in bat species, as 4 of the 5 tropical studies are for mammal and bats were the only nectar consuming mammals sampled in our data. Interpretation of the increases for temperate studies is less clear, as temperate studies included many bird communities with a variety of nectar consumers. Similarly, decreasing trends for vertebrate consumption in marine and tropical studies are not obviously consistent with known changes in those contexts and warrant further examination.", "blue")`

### Study heterogeneity

@@

<!-- huge differences between different taxa and the kind of disturbance/land use change -->

```{r changeScenarios, echo = FALSE, message = FALSE, fig.show = 'hold', fig.cap= "Comparison of the distribution of time series durations from each type of change to the overall distribution of time series durations, with no richness change or turnover and functional change in yellow, no turnover but richness and functional change in green, no change in blue, and turnover, richness and functional change studies in orange."}
# 
# knitr::include_graphics(here::here("figures/duration_hist.jpeg"))
# 
# knitr::include_graphics(here::here("figures/functional_change_table.png"))
```

### Areas of incongruence

`r colorize("The addition of functional trait data illuminates a few key areas where a synthesis approach is incongruous with system-specific studies or expectations based on anthropegenic impacts. For example, we found no evidence of general trends in the *CWM* for body mass or for studies aggregated by their characteristics despite predictions that mean body mass is decreasing as a result of climate change impacts and megafaunal loss ", "blue")`[@sheridan2011] `r colorize(", and downgrading documented empirically and experimentally in multiple taxa ", "blue")`[@caruso2014; @forster2012; @huss2019; @tseng2018].
`r colorize("Still, climate change impacts are predicted to be most extreme for ectotherms and impact intraspecific traits first before species are lost, neither of which are represented by our data. Further, many of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing", "blue")` [@fritz2009].
`r colorize("Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened", "blue")` [@ripple2015].

`r colorize("Results for protected areas were also surprising given the assumption that protection insulates communities from functional degradation. The only difference we found between protected and unprotected areas was significant turnover occurring outside protected areas and no significant turnover inside protected areas. Rather than the result of protection itself, this is likely due to the fact that the majority of studies were form terrestrial, temperate studies where turnover is known to be lower ", "blue")`[@blowes2019].
`r colorize("While things may generally be more static within protect areas, the functional dimensions of protected communities fared no better or worse on average than their non-protected counter parts. Looking at individual studies, there is a mix of both losses and gains in almost all functional diversity and composition metrics across communities found in protected areas.", "blue")`

>>>>

`r colorize("The expectation for functional diversity change across communities is not obvious from past work, and may or may not follow taxonomic trends", "blue")` [@dornelas2023].
While `r colorize("loss of functional diversity", "blue")` is frequently cited as one of the most pressing concerns of the anthropocene [@cardinale2012; @dirzo2014; @young2016], `r colorize("functional diversity may be maintained even when species are lost from a community", "blue")` [@daz2001; @petchey2009].
Forecasts of functional loss range from negligible [@gallagher2013] to dire [@petchey2002; @pimiento2020].
And while some observed trends show significant functional loss [@flynn2009] others document no loss even in some of the most heavily impacted communities [@edwards2013; @larsen2018; @matuoka2020].
On paleoecological time scales, `r colorize("the functional space", "blue")` shows mixed responses to environmental change and extinction events [@jackson2015; @dornelas2018], with significant impacts of species extinctions on functional diversity in some taxa and not others [@pimiento2017].
For some time periods, functional diversity appears to be maintained for substantial portions of geological time [@hedberg2022].
`r colorize("Contemporary, broad-scale examinations of functional change are limited to only a few taxa-focused studies, but show for example functional richness increases for both North American birds", "blue")` [@barnagaud2017; @jarzyna2016] `r colorize("and ray-finned fishes, sharks, and rays", "blue")` [@trindade-santos2020].

We have stronger expectations for changes in the prevalence of some individual traits.
For example, animal body size is expected to decrease as a result of climate change, a phenomena that has been documented in multiple taxa empirically and experimentally [@caruso2014; @forster2012; @huss2019; @sheridan2011; @tseng2018].
There is some evidence that this holds true for birds [@vanbuskirk2010], but the picture is likely more complicated for mammals, where urbanization may actually lead to larger body sizes due to novel food sources [@hantak2021], even as megafaunal loss leads to on average smaller community body size.
For dietary traits, recent work documenting insect declines (citation) points to potentially significant negative impacts on insectivorous animals [@canaday1996; @sekercioglu2002; @tallamy2021].
Predicted extinctions based on species-level vulnerability point to further dietary shifts, favoring increases in invertivorous species [@cooke2019].
Some systems also show significant shifts in the prevalence of different kinds of foragers in birds, for example loss of arboreal foragers in agricultural systems [@bain2020], and loss of neotropical understory foragers even in protect areas [@pollock2022].

Here we `r colorize("leverage ongoing efforts to assemble functional trait data and recent computational advances to", "blue")` perform the first multi-taxa, multi-realm assessment of functional diversity and composition change through time.
We focus on mammal and bird species as `r colorize("a subset of the world's biodiversity of particular conservation concern that is heavily impacted by anthropogenic change.", "blue")`While examining trends in plants, invertebrates, and other vertebrate species is of equal interest, trait data for those taxa raise additional challenges such as limited and biased species coverage [@fitzjohn2014], a lack of accepted species-level means, and differences in the types of traits collected.
To ensure comparability across taxa in trait type and data quality we therefore focus on mammals and birds.
`r colorize("We include body mass, dietary, foraging and other behavioral traits", "blue")` that were intentionally selected to be representative of a species' Eltonian niche, thereby summarizing the functional role they play in the community [@wilman2014].
An initial assessment of amphibian trends is included in the supplement, but excluded from general trend assessment here due to limited geographic coverage.

We assess thousands of mammal and bird functional diversity time series to determine whether or not `r colorize("the addition of functional trait data gives a clarifying picture of biodiversity change across communities. We present a few areas of change consensus across communities, and even more scenarios where the relationship between functional and species change are unexplained and warrant further theoretical and experimental examination. We present results for three different levels: general trends across communities, trends for communities with similar characteristics (taxa, biome, realm, protection status), and trends for individual studies.", "blue")` We further evaluate evidence for key predictions of individual trait prevalence from the literature, including changes in mean body size, dietary traits, and foraging strata.\`

@@

```{r traitTable, echo=FALSE}
tibble(Trait = c(rep("Diet", 10), rep("Foraging Strata", 7), rep("Pelagic Specialist", 2), rep("Nocturnal", 2), rep("Crepuscular", 2), rep("Diurnal", 2), "Body Mass"),
       Category = c("Invertebrate", "Mammals and Birds", "Reptiles", "Fish", "Unknown Vertebrates", "Scavenging", "Fruit", "Nectar", "Seeds", "Other Plant", 
                    "Below water surface", "water surface", "ground", "understory", "> 2m, below canopy", "canopy", "aerial",
                    rep(c("yes", "no"), 4),
                    "-"),
       Taxa = c(rep("Bird and Mammal", 10), rep("Bird", 7), rep("Bird", 2), rep("Bird and Mammal", 2), rep("Mammal", 4), "Bird and Mammal"),
       `Data Type` = c(rep("percentage consumed", 10), rep("percentage of use", 7), rep("binary", 8), "continous, in grams")) %>%
  kable(format = "latex", caption = "Description of the traits included in the analysis broken down by categories at data type.") %>%
  collapse_rows(columns = 1:4) %>%
  column_spec(3:4, width = "3cm") %>%
  kable_paper()
```

@@

Functional diversity metrics were calculated using the *dbFD* function from the *FD* R package [@laliberté2010].
Here we report functional richness (*FRic*), functional evenness (*FEve*), and functional divergence (*FDiv*) which together describe three complementary characteristics of the functional space [@mason2005; @hillebrand2009].
*FRic* assesses the volume of the trait space occupied by species in the community, with higher values indicating communities with species of more extreme trait values.
*FEve* describes how species are distributed across the trait space and how abundance is distributed across species.
Higher values of *FEve* indicate more even spacing of species in the trait space and individuals across species.
*FDiv* measures the degree to which species and their abundances maximize differences in the functional space.
Higher values of *FDiv* therefore correspond to communities where many highly abundant species are on the edges of the trait space.

We also calculated the community-weighted mean (*CWM*) of all continuous traits to examine `r colorize("turnover", "blue")` in the distribution of each trait.
`r colorize("Wholesale shifts in the trait space due to changes of trait means could occur even while the shape of the multidimensional trait space, as defined by functional richness, evenness, and divergence, is maintained. CWM's are therefore a way to assess whether or not that turnover is occurring and what the nature of the shift may be. Hereafter we refer to results for functional metrics in two groups: functional diversity metrics (FRic, FEve, FDiv) and composition metrics (trait CWM's).", "blue")`

@@

## Evidence of consensus

@@

`r colorize("We found evidence of functional richness loss in mammal studies, with a loss trend significantly different from zero and significantly different from the trend for birds, indicating a potential loss of functional capacity in these communities. This result is consistent with previous work linking anthropogenic drivers such as habitat loss, poaching, and human modification and development to loss of functional diversity in mammal communities across the globe", "blue")` [@brodie2021; @cox2022; @li2022].
`r colorize("However we did not find trends outside the random expectation for species loss, in contrast to expectations for multiple continents based on IUCN risk categories", "blue")` [@brodie2021].
`r colorize("In general, functional capacity of mammals showed greater declines that bird communities.", "blue")`

## Areas of incongruence

`r colorize("The addition of functional trait data illuminated multiple areas where a synthesis approach is incongruous with system-specific studies or simply unexpected. For example, results for protected areas were surprising given the assumption that protection insulates communities from functional degradation. The only difference we found between protected and unprotected areas was significant turnover occurring outside protected areas and no significant turnover inside protected areas. Rather than the result of protection itself, this is likely due to the fact that the majority of studies were from terrestrial, temperate studies where turnover is known to be lower ", "blue")`[@blowes2019].
`r colorize("While things may generally be more static within protect areas, the functional dimensions of protected communities fared no better or worse on average than their non-protected counter parts. Looking at individual studies, there is a mix of both losses and gains in almost all functional diversity and composition metrics across communities found in protected areas.", "blue")`

`r colorize("We also found evidence of increasing evenness for marine studies, which was not predicted by previous work in marine systems. The empirical link between changes in evenness and ecosystem process is the most poorly studied of the metrics, however theory indicates that increased evenness improves function and stability ", "blue")` [@filstrup2019], `r colorize("as the dominance of one or a few species makes function sensitive to the ability of those species to respond to environmental change ", "blue")`[@hillebrand2008; @hillebrand2009].
`r colorize("Further, there is some evidence that maintaining evenness is important for supporting multifunctionality in communities ", "blue")`[@bagousse-pinguet2019].
`r colorize("An increase in evenness is therefore a positive indicator for these communities, despite the fact that marine bird and mammal communities are some of the most threatened communities in the world. Still, this result may be a function of data limitations as the majority of marine studies are of seabird communities (7 of 9), so the general trend may be an early indicator of pay off from investments in sea bird recovery, which is also indicated in recent species-based work ", "blue")`[@spatz2023].

Our results were also inconsistent with all predictions for changes in trait prevalence, including foraging strata, body size, and dietary traits.
`r colorize("We found no changes in the prevalence of different foraging strategies, despite documented losses of understory birds in the neotropics and some evidence that agricultural incursion particularly threatens arboreal species. These shifts may therefore be the result of specific contexts, and not generalizable to bird communities across the globe. We found no evidence of general trends in the *CWM* for body mass for birds or mammals, indicating that either body size is not changing significantly due to climate change, opposing pressures such as urbanization are overshadowing climate change impacts, or more likely current shifts are happening at an intraspecific level and and have not yet propogated to the species losses that would be captured by our data. Further, many of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing", "blue")` [@fritz2009].
`r colorize("Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened", "blue")` [@ripple2015].

Rather than declines in insectivory and increases in invertivory predicted by previous work, we instead found changes in the degree of consumption of fish, nectar, and vertebrates.
`r colorize("Increases in fish consumption for bird and temperate studies may be the product of increased availability of easy fish from commercial fishing operations or the stocking of inland waterways for sport fishing. However, fisheries also have negative impacts on waterbirds through competition for fish or bird mortality due to by catch, making these results suprising. Decreases in nectar consumption for tropical studies is likely due to declines in bat species, as 4 of the 5 tropical studies are for mammal and bats were the only nectar consuming mammals sampled in our data. Interpretation of the increases for temperate studies is less clear, as temperate studies included many bird communities with a variety of nectar consumers. Similarly, decreasing trends for vertebrate consumption in marine and tropical studies are not obviously consistent with known changes in those contexts and warrant further examination.", "blue")`

## Study heterogeneity

@@
